A decade of lost precision: Introducing a general framework for image feature matching without geometric constraints

### 2014-04-11 text/journal/paper.tex
Corrected introduction and added citation

\begin{document}
\maketitle
%
\begin{abstract}
Solutions in computer vision involving the matching of local image 
features frequently use \emph{Ratio-Match} as introduced by Lowe and 
others, but is this really the best approach? We formalize the 
theoretical foundation of \emph{Ratio-Match} and propose a general 
framework encompassing \emph{Ratio-Match} and five other matching 
methods. Using the framework we establish a theoretical order of 
performance in terms of precision and recall between the methods and 
prove that \emph{Ratio-Match} is consistently equalled or outperformed 
by all other methods of the framework.

@@

A plethora of solutions have combined ratio match with various geometric 
constraints to improve matching. These constraints are
based on assumptions regarding the transformation between the query and 
target images. At the stricter end we have epipolar constraints assuming
that the images can be tied by a homography \cite{torr2000mlesac}, 
\cite{chum2005matching} and angular constraints assuming the 
correspondences are angled similarly \cite{kim2008efficient}, 
\cite{schmid1997local}, \cite{leordeanu2005spectral}. Instead of 
modelling the relationship between two images as an epipolar relation, 
we can try to allow for more lenient transformation with more 
sophisticated models.  Torresani et al.~\cite{torresani2008feature} 
attempted to model these transformations using graph matching similar to 
\cite{deriche1994robust}. Others define image regions and reject or 
accept correspondences based on the regions they connect 
\cite{cho2009feature}, \cite{wu2011robust}.

Any matching method relying on geometric constraints is restrained by an
inherent assumptions about the geometric relationship between the two 
images. Broad assumptions such as the epipolar constraint only apply in 
simple image transformations. For more complex transformations we need 
models suitable for each particular case and hence limit the algorithm 
to the subset of images that fit the model.  In the case of object 
recognition for example, the transformations from one scene to another 
often feature a change
in perspective, background and sometimes variations within the object 
itself.  A person can change pose, a car model can have different 
configurations, a flower can bloom etc.

When matching these instances we are forced to either create a 
sophisticated model that represents the variables of transformation 
within the object, or alternatively find correspondences using an 
algorithm with no inherent geometric assumptions. In addition the 
advantage of purely descriptor based methods is motivated by the fact 
that any geometric method acts as a filter on a given set of 
correspondences.  If this set contains less incorrect correspondences, 
the final set can be calculated faster and more accurately.

@@

@@

\section{Matching Methods}
\label{S:MatchingMethods}
%
\subsection{Ratio-Match}
%
\emph{Ratio-Match} as introduced by Deriche et al.
\cite{deriche1994robust} and later used by 
Baumberg~\cite{baumberg2000reliable} and Lowe~\cite{lowe2004sift} makes 
use of a \emph{uniqueness ratio}. For each feature in the \emph{query 
image} we find a possible match by identifying the nearest neighbor and 
returning this match only if the \emph{uniqueness ratio} is above a 
given threshold, $\tau$. Given $f_q$, a feature from the \emph{query 
image} and $F_{target}$, a set of features from the \emph{target image}, 
the nearest neighbor of $f_q$ in $F_{target}$ is:
\begin{equation*}
    f_{nn} = \argmin_{f_t \in F_{target}} d(f_q, f_t)
\end{equation*}
Here, $d(f_i, f_j)$ is the distance between the feature descriptors.  
With SIFT and SURF this is the euclidian distance of the feature vectors
\cite{lowe2004sift} \cite{bay2006surf} where as BRIEF, BRISK and FREAK 
measure the distance as the hamming distance \cite{leutenegger2011brisk} 
\cite{calonder2010brief} \cite{alahi2012freak}.  If we let $f_p$ and 
$f_b$ be the nearest and second nearest neighbor of a feature, $f_q$, 
the \emph{uniqueness ratio} is defined as follows:
\begin{equation*}
    r = d(f_q, f_p) / d(f_q, f_b)
\end{equation*}
Because $f_p$ is the nearest neighbor and $f_b$ the second nearest 
neighbor we know that $d(f_q, f_p) \leq d(f_q, f_b)$ and since $d(f_i, 
f_j) \geq 0, \forall f_i,f_j$, the uniqueness ratio is bounded by $0$ 
and $1$, both inclusive. For the degenerate case where $d(f_q, f_b) = 0$
we let the \emph{uniqueness ratio} be $0$.

@@

\begin{itemize}
\item{Let $f_q$ be any feature point in the \emph{query image} that we 
    are currently trying to match}
\item{Let $F$ be a set of features. $F_{target}$ for example is the set 
    of features from the \emph{target image}}
\item{Let $\tau \in [0 \ldots 1]$ be a threshold used to decide whether 
    to keep a match or discard it. We keep a match only if is has a 
\emph{uniqueness ratio} lower than $\tau$}
\item{Let the \emph{proposed match} be the nearest neighbor of $f_q$
    picked from a set of feature points that we will call the 
\emph{proposal set} which does not include $f_q$}
\item{Let the \emph{baseline match} be the nearest neighbor of $f_q$ 
    picked from a set of feature points that we will call the 
\emph{baseline set} which neither contains the \emph{proposed match} nor 
$f_q$}
\end{itemize}

\begin{figure}[b]
\centering
\includegraphics[width=0.85\columnwidth]{images/overview}
\caption{Graphical representation of the \emph{Baseline} and 
\emph{Proposal Set} for different methods in the proposed framework}
\label{fig:overview}
\end{figure}

In the case of \emph{Ratio-Match} \cite{lowe2004sift} the \emph{proposal 
set} consists of all the feature points in the target image.  Similarly 
the \emph{baseline set} consists of all the feature points in the target 
image except the \emph{proposed} match. For \emph{Ratio-Match} this 
practically means that given a feature, $f_q$, the \emph{proposed match} 
and the \emph{baseline match} are the two nearest neighbors in the set 
of feature points in the \emph{target image}.

Brown et al.\ extends \emph{Ratio-Match} for the case of $n$ target 
images by letting the baseline be the average of the distance to the 
second nearest neighbors for all $n$ target images 
\cite{brown2005multi}.  ``\emph{[W]e can improve outlier rejection}'' 
they write ``\emph{by using information from all of the images (rather 
than just the two being matched) This improves the matching 
precision}''.  The idea of improving matching precision by expanding the 
knowledge used to reject outliers is the central tenet in this paper.

%
\subsection{A framework of matching methods}
%
\begin{table}[t]
\caption{Calculating the uniqueness ratio using different sources for 
    the \emph{proposal set} and \emph{baseline set} for the case of two 
images.}
\label{table:distinctness}
	\centering
%	\small
\bgroup
\def\arraystretch{1.5}
    \begin{tabular}{l l | l l}
        %
        %&  \multicolumn{1}{l}{Target}
        %& \multicolumn{1}{l}{Target and Query} \\
\multicolumn{2}{r}{} & \multicolumn{2}{c}{\emph{Proposal Set}} \\
                     & & Target & Target and Query \\
        \cline{2-4}
        \multirow{3}{*}{\rotatebox[origin=c]{90}{\emph{Baseline Set}}} &
  Query            & \emph{Self-Match}   & \emph{Self-Match-Ext} \\
& Target           & \emph{Ratio-Match}   & \emph{Ratio-Match-Ext} \\
& Target and Query & \emph{Mirror-Match} & \emph{Mirror-Match-Ext} \\
        \cline{2-4}
\end{tabular}
\egroup
\end{table}

\emph{Ratio-Match} is only one instance of a framework of algorithms we can 
construct by different combinations of \emph{proposed} and 
\emph{baseline sets}. As shown in Table~\ref{table:distinctness} for the 
case of two images the \emph{proposal set} can be constructed by either 
looking only at the nearest neighbors in the target image, or 
alternatively by looking at the nearest neighbors in both the query and 
target image. The \emph{baseline set} can either be the query image 
itself, the target image or the combination of the two.  Seen in this 
light, the matching algorithm \emph{Ratio-Match} draws both its 
\emph{proposal set} and \emph{baseline set} from the target image. In 
turn \emph{Mirror-Match-Ext} as introduced by the authors in 
\cite{arnfred2013mirror}, works by constructing the \emph{proposal set} 
and \emph{baseline set} by combining features from both the \emph{query} 
and \emph{target image}. Alternatively \emph{Self-Match} looks only to 
the query image for the \emph{baseline set} but constructs the 
\emph{proposal set} from the target image like \emph{Ratio-Match}. In 
Figure~\ref{fig:overview} we illustrate how each method is derrived.

\begin{algorithm}[htb]
\caption{Generalized matching algorithm for two images}
\label{alg-gen}
\begin{algorithmic}
    \Require $I_{query}, I_{target}$ : images, $\tau \in [0,1]$
    \State $F_{query} = get\_features(I_{query})$
    \State $F_{target} = get\_features(I_{target})$
\State $F_{proposal} = get\_proposal\_set(F_{query}, F_{target})$
\State $F_{baseline} = get\_baseline\_set(F_{query}, F_{target})$
\State $M = \varnothing$
\ForAll{$f_q \in F_{query}$}
    \State $f_p \gets getNearestNeighbo\text{r}(f_q, F_{proposal} \setminus 
    \left\{f_q\right\})$
    \State $f_b \gets getNearestNeighbo\text{r}(f_q, F_{baseline} \setminus 
    \left\{f_q, f_p\right\})$
    \State $ratio \gets distance(f_q, f_p) / distance(f_q, f_b)$
    \If{$(ratio < \tau) \wedge (f_p \in F_{target})$}
        \State $matches \gets matches \cup \left(f_q, f_p\right)$
	\EndIf
\EndFor
\Return $M$
\end{algorithmic}
\end{algorithm}

In algorithm~\ref{alg-gen} we show the generalized matching framework 
where the \emph{proposal set} and \emph{baseline set} are defined 
according to Table~\ref{table:distinctness}.  In practice for i.e.\ 
\emph{Ratio-Match} we can obtain $f_p$ and $f_b$ by finding the 2 
nearest neighbors of $F_{target}$. The other algorithms can be optimized 
similarly. In Figure~\ref{fig:matching} we illustrate the structure of 
this algorithm.

@@

For $F_{query} = \left\{f_{q_1} \ldots f_{q_n}\right\}$ and $F_{target} 
= \left\{f_{t_1} \ldots f_{t_m}\right\}$.  We can obtain $F_{proposed}$ 
and $F_{baseline}$ from these sets according to the algorithm used (see 
Table~\ref{table:distinctness}). $r = \text{r}(f_{q}, F_{proposed}, 
F_{baseline})$ is the ratio of query feature $f_{q}$ given 
$F_{proposed}$ and $F_{baseline}$.  If $f_{p}$ is the nearest neighbor 
of $f_{q}$ in $F_{proposed}$ and $f_{b}$ is the nearest neighbor of 
$f_{q}$ in $F_{baseline} \setminus \left\{f_{p}\right\}$ then the 
generalized \emph{uniqueness} ratio is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \frac{d(f_{q}, f_{p})}{d(f_{q}, f_{b})}
\end{align*}
We let $K$ be the number of possible true correspondences \emph{query} 
and \emph{target image}. The \emph{precision} and \emph{recall} are then 
defined as:
\begin{align*}
    \textrm{Recall} &= \frac{\#\textrm{Correct Matches}}{K} \\
    \textrm{Precision} &= \frac{\#\textrm{Correct 
    Matches}}{\#\textrm{Correct Matches} + \#\textrm{Incorrect Matches}}
\end{align*}
Given $f_{nn} \in F_{target}$ defined as the nearest neighbor of any 
$f_q \in F_{query}$ then we define the set of features $F_{true}$ as 
$\left\{ f_{q} \in F_{query} \mid f_{nn} \text{ is a true correspondence 
of } f_{q} \right\}$ and $F_{false}$ as $F_{query} \setminus F_{true}$.  
We can the define the number of correct and incorrect matches as 
follows:
\begin{align*}
    \textrm{keep\_match}(f_{q}) &= \twopartdef{ 1 }{\text{r}(f_{q}, 
    F_{proposed}, F_{baseline}) <
    \tau}{0}{\textrm{otherwise}} \\
    \#\textrm{Correct Matches} &= \sum_{f_{q} \in F_{true}} 
    \textrm{keep\_match}(f_{q})\\
    %\mathbbm{1}_{\left\{ \right\}} \\
    \#\textrm{Incorrect Matches} &= \sum_{f_{q} \in F_{false}}
    \textrm{keep\_match}(f_{q})
    %\mathbbm{1}_{\left\{ r < \tau\right\}}
\end{align*}

@@

Under three assumptions that largely reflects the performance of 
matching feature points in practice we can theoretically compare the 
performance of the different algorithms shown in 
Table~\ref{table:distinctness}.

@@

\begin{figure*}[t!]
    \begin{subfigure}[t]{1\textwidth}
        \centering
        \includegraphics[width=1\columnwidth]{images/results_descriptors_imageset1}
        \caption{Keypoint /Descriptor combinations evaluated on 3D 
        Objects}
        \label{fig:descriptors}
    \end{subfigure}%

    \caption{Keypoint / Descriptor combinations measured on 15 pairs of 
    photos of 3D objects taken 15 degrees apart. K = Keypoint, D = 
Descriptor\.}
\end{figure*}

>>>>

\begin{document}
\maketitle
%
\begin{abstract}
Solutions in computer vision involving the matching of local image 
features frequently use \emph{Ratio-Match} as introduced by Lowe and 
others, but is this really the best approach? We formalize the 
theoretical foundation of \emph{Ratio-Match} and propose a general 
framework encompassing \emph{Ratio-Match} and three other matching 
methods. Using the framework we establish a theoretical order of 
performance in terms of precision and recall between the methods and 
prove that \emph{Ratio-Match} is consistently equalled or outperformed 
by all other methods of the framework.

@@

A plethora of solutions have combined ratio match with various geometric 
constraints to improve matching. These constraints are based on 
assumptions regarding the transformation between the query and target 
images. At the stricter end we have epipolar constraints assuming
that the images can be tied by a homography \cite{torr2000mlesac}, 
\cite{chum2005matching} and angular constraints assuming the 
correspondences are angled similarly \cite{kim2008efficient}, 
\cite{schmid1997local}. Often these approaches are made computationally 
feasible by modelling the case of feature correspondences as an instance 
of graph matching where each feature is a vertex and edge values 
correspond to a geometric relation between two features. In this case 
approximate graph matching algorithms can be used to efficiently 
establish an isomorphism between the graph of features in two images 
\cite{leordeanu2005spectral}, \cite{torresani2008feature}, 
\cite{yarkony2010covering}. Finally others define image regions and 
reject or accept correspondences based on the regions they connect 
\cite{cho2009feature}, \cite{wu2011robust}.

Any matching method relying on geometric constraints is restrained by an
inherent assumptions about the geometric relationship between the two 
images. Broad assumptions such as angular constraints often used in 
graph matching only apply to simple image transformations. For more 
complex transformations we need models suitable for each particular case 
and hence limit the algorithm to the subset of images that fit the 
model. In the case of object recognition for example, the 
transformations from one scene to another often feature a change in 
perspective, background and sometimes variations within the object 
itself. In most circumstances this makes these methods unsuited for 
epipolar or angular constraints.

If we intent to match these instances using geometric features we are 
forced to used a sophisticated model that reflects the variations in 
which we find these objects, an in the process lose generality. For this 
reason general descriptor based algorithms are crucial for 
correspondence applications calling for generality. In addition the 
advantage of purely descriptor based methods is motivated by the fact 
that any geometric method acts as a filter on a given set of 
correspondences.  If this set contains less incorrect correspondences, 
the final set can be calculated faster and more accurately.

@@

In the author's last paper we introduced \emph{Mirror Match} and 
\emph{Mirror-Match with Clustering}, two algorithms that we showed to 
perform better than the state of the art. The novel contributions of 
this paper consists of two new algorithms completing a framework of four 
related matching algorithms as well as the theoretical foundation which 
enables us to understand why \emph{Mirror Match} performs better than 
\emph{Ratio Match} in the first place.

@@

\section{Matching Methods}
\label{S:MatchingMethods}
%
\subsection{Ratio-Match}
%
\emph{Ratio-Match} as introduced by Deriche et al.
\cite{deriche1994robust} and later used by 
Baumberg~\cite{baumberg2000reliable} and Lowe~\cite{lowe2004sift} makes 
use of a \emph{uniqueness ratio}. For each feature in the \emph{query 
image} we find a possible match by identifying the nearest neighbor and 
returning this match only if the \emph{uniqueness ratio} is above a 
given threshold, $\tau$. Given $f_q$, a feature from the \emph{query 
image} and $F_{target}$, a set of features from the \emph{target image}, 
the nearest neighbor of $f_q$ in $F_{target}$ is:
\begin{equation*}
    \argmin_{f_t \in F_{target}} d(f_q, f_t)
\end{equation*}
Here, $d(f_i, f_j)$ is the distance between the feature descriptors.  
With SIFT and SURF this is the euclidian distance of the feature vectors
\cite{lowe2004sift} \cite{bay2006surf} where as BRIEF, BRISK and FREAK 
measure the distance as the hamming distance \cite{leutenegger2011brisk} 
\cite{calonder2010brief} \cite{alahi2012freak}.  If we let $f_p$ and 
$f_b$ be the nearest and second nearest neighbor of a feature, $f_q$, 
the \emph{uniqueness ratio} is defined as follows:
\begin{equation*}
    r = d(f_q, f_p) / d(f_q, f_b)
\end{equation*}
Because $f_p$ is the nearest neighbor and $f_b$ the second nearest 
neighbor we know that $d(f_q, f_p) \leq d(f_q, f_b)$ and since $d(f_i, 
f_j) \geq 0, \forall f_i,f_j$, the uniqueness ratio is bounded by $0$ 
and $1$, both inclusive. For the degenerate case where $d(f_q, f_b) = 0$
we let the \emph{uniqueness ratio} be $0$.

@@

\begin{itemize}
\item{Let $f_q$ be any feature point in the \emph{query image} that we 
    are currently trying to match}
\item{Let $F$ be a set of features. $F_{target}$ for example is the set 
    of features from the \emph{target image}}
\item{Let $\tau \in [0 \ldots 1]$ be a threshold used to decide whether 
    to keep a match or discard it. We keep a match only if is has a 
\emph{uniqueness ratio} lower than $\tau$}
\item{Let the \emph{proposed match} be the nearest neighbor of the query 
    feature $f_q$ picked from a set of feature points that we will call 
the \emph{proposal set} which does not contain the query feature itself}
\item{Let the \emph{baseline match} be the nearest neighbor of the query 
    feature $f_q$ picked from a set of feature points that we will call 
the \emph{baseline set} which neither contains the \emph{proposed match} 
nor the query feature}
\end{itemize}

%
\subsection{A framework of matching methods}
%

\begin{figure}[t]
\centering
\includegraphics[width=0.85\columnwidth]{images/overview}
\caption{Graphical representation of the \emph{Baseline} and 
\emph{Proposal Set} for different methods in the proposed framework}
\label{fig:overview}
\end{figure}

Given two images we can construct the proposal and baseline sets in six 
different ways as illustrated in Figure~\ref{fig:overview}. In the case 
of \emph{Ratio-Match} \cite{lowe2004sift} the \emph{proposal set} 
consists of all the feature points in the target image.  Similarly the 
\emph{baseline set} consists of all the feature points in the target 
image except the \emph{proposed} match. For \emph{Ratio-Match} this 
practically means that given a query feature, $f_q$, the \emph{proposed 
match} and the \emph{baseline match} are the two nearest neighbors in 
the set of feature points in the \emph{target image} and the ratio is 
decided uniquely based on the \emph{target image}. In turn 
\emph{Mirror-Match-Ext} as introduced by the authors in 
\cite{arnfred2013mirror}, works by constructing the \emph{proposal set} 
and \emph{baseline set} by combining features from both the \emph{query} 
and \emph{target image}.

\begin{algorithm}[htb]
\caption{Generalized matching algorithm for two images}
\label{alg-gen}
\begin{algorithmic}
    \Require $I_{query}, I_{target}$ : images, $\tau \in [0,1]$
    \State $F_{query} = get\_features(I_{query})$
    \State $F_{target} = get\_features(I_{target})$
\State $F_{proposal-all} = get\_proposal\_features(F_{query}, 
F_{target})$
\State $F_{baseline-all} = get\_baseline\_features(F_{query}, 
F_{target})$
\State $M = \varnothing$
\ForAll{$f_q \in F_{query}$}
    \State $F_{proposal} = F_{proposal-all} \setminus 
    \left\{f_q\right\}$
    \State $F_{baseline} = F_{baseline-all} \setminus \left\{f_q, 
    f_p\right\}$
    \State $f_p \gets getNearestNeighbo\text{r}(f_q, F_{proposal})$
    \State $f_b \gets getNearestNeighbo\text{r}(f_q, F_{baseline})$
    \State $ratio \gets distance(f_q, f_p) / distance(f_q, f_b)$
    \If{$(ratio < \tau) \wedge (f_p \in F_{target})$}
        \State $matches \gets matches \cup \left(f_q, f_p\right)$
	\EndIf
\EndFor
\Return $M$
\end{algorithmic}
\end{algorithm}

In algorithm~\ref{alg-gen} we show the generalized matching framework 
where the \emph{proposal set} and \emph{baseline set} are defined 
according to Figure~\ref{fig:overview}.  In practice for i.e.\ 
\emph{Ratio-Match} we can obtain $f_p$ and $f_b$ by finding the 2 
nearest neighbors of $F_{target}$. The other algorithms can be optimized 
similarly. In Figure~\ref{fig:matching} we illustrate the structure of 
this algorithm.

@@

For $F_{query} = \left\{f_{q_1} \ldots f_{q_n}\right\}$ and $F_{target} 
= \left\{f_{t_1} \ldots f_{t_m}\right\}$.  We can obtain $F_{proposed}$ 
and $F_{baseline}$ from these sets according to the algorithm used (see 
Figure~\ref{fig:overview}). $r = \text{r}(f_{q}, F_{proposed}, 
F_{baseline})$ is the ratio of query feature $f_{q}$ given 
$F_{proposed}$ and $F_{baseline}$.  If $f_{p}$ is the nearest neighbor 
of $f_{q}$ in $F_{proposed}$ and $f_{b}$ is the nearest neighbor of 
$f_{q}$ in $F_{baseline} \setminus \left\{f_{p}\right\}$ then the 
generalized \emph{uniqueness} ratio is:
\begin{align*}
    r &= \text{r}(f_{q}, F_{proposed}, F_{baseline}) \\
        &= \frac{d(f_{q}, f_{p})}{d(f_{q}, f_{b})}
\end{align*}
We let $K$ be the number of possible true correspondences \emph{query} 
and \emph{target image}. The \emph{precision} and \emph{recall} are then 
defined as:
\begin{align*}
    \textrm{Recall} &= \frac{\#\textrm{Correct Matches}}{K} \\
    \textrm{Precision} &= \frac{\#\textrm{Correct 
    Matches}}{\#\textrm{Correct Matches} + \#\textrm{Incorrect Matches}}
\end{align*}
Given $f_{nn} \in F_{target}$ defined as the nearest neighbor of any 
$f_q \in F_{query}$ then we define the set of features $F_{true}$ as 
$\left\{ f_{q} \in F_{query} \mid f_{nn} \text{ is a true correspondence 
of } f_{q} \right\}$ and $F_{false}$ as $F_{query} \setminus F_{true}$.  
We can the define the number of correct and incorrect matches as 
follows:
\begin{align*}
    \textrm{keep\_match}(f_{q}) &= \twopartdef{ 1 }{\text{r}(f_{q}, 
    F_{proposed}, F_{baseline}) <
    \tau}{0}{\textrm{otherwise}} \\
    \#\textrm{Correct Matches} &= \sum_{f_{q} \in F_{true}} 
    \textrm{keep\_match}(f_{q})\\
    %\mathbbm{1}_{\left\{ \right\}} \\
    \#\textrm{Incorrect Matches} &= \sum_{f_{q} \in F_{false}}
    \textrm{keep\_match}(f_{q})
    %\mathbbm{1}_{\left\{ r < \tau\right\}}
\end{align*}

@@

Under three assumptions that largely reflects the performance of 
matching feature points in practice we can theoretically compare the 
performance of the different algorithms shown in 
Figure~\ref{fig:overview}.

@@

\begin{figure*}[t]
    photos of 3D objects taken 15 degrees apart. K = Keypoint, D = 
Descriptor\.}
    \caption{Keypoint / Descriptor combinations measured on 15 pairs of 
    photos of 3D objects taken 15 degrees apart. K = Keypoint, D = 
Descriptor.}
    \label{fig:descriptors}
\end{figure*}
